Wealth dynamics in the presence of network structure and primitive cooperation

### 2019-07-22 manuscript/draftcssa19.tex

\end{abstract}
\section{Introduction}
In recent years, the concern and debate regarding wealth inequality and socioeconomic mobility have been one of the few unifying issues dominating the extremely polarized public spheres of the Global North. While economic and political inequality used to be discussed in heterodox economics circles, the contentious discussions on the topic within mainstream economics since the publication of Piketty's book~\cite{piketty2017capital} suggest a lack of consensus about basic foundational questions like the origins and persistence of economic inequality. Not surprisingly, traditional theories and tools of macro and micro economics are now being diagnosed for their limitations. Simultaneously, insights from related disciplines, along with novel models for scientific inquiry not usually associated with traditional econometrics, are being taken more seriously. The work presented here shares this spirit by integrating substantive ideas from anthropology, economic sociology~\cite{granovetter2017society} and urban sociology~\cite{sampson2012great} and using modeling approaches from computational social science~\cite{hedstrom2018}, computational economics~\cite{tesfatsion} and analytical sociology~\cite{hedstrom2011oxford} to explore the origins of inequality in a simple model of wealth dynamics in the presence of social structures.    
% SD: citation for "anthropology," the first element of the last sentence's list? Grief and/or White?

@@

These models were inspired by our search for analogs of the grid world agent-based models (ABMs) of Friesen and Mudigonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. The original model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

The original model drew its inspiration from historical sociology, in Katz's influential study of middle of 19th century Hamilton, Canada~\cite{katz2013people}. Retaining this original motivation, we draw additional inspiration from economic sociology in the work of Granovetter~\cite{granovetter2005}; urban sociology, in the work of Sampson~\cite{sampson2002} and others; and in anthropology, in the work on cooperation in small to medium scale societies~\cite{avner1994,white2011kinship}, and others. The geographic and economic scale of the systems and the nature of social actors and time period of interest are all very different from the ones used to develop models of representative agents in macroeconomics~\cite{benhabib2018}, making the similarity between macroeconomic wealth dynamics models and our models not comparable without further justification. Elaborating on the interplay of conceptual and methodological ideas among these disciplines is beyond the scope of this article. Instead, we anchor this work in economic sociology and revisit the insights from the above disciplines in the article's concluding section.

The important role played by social structures or non-economic structures in determining economic outcomes of individuals in a society is not in doubt~\cite{granovetter2005, jackson_rev2017}. Still, in the absence of a unifying foundation for sociology and economics, the full impact of this two-way interpenetration of economic and social structures is demonstrated only on a case-by-case basis. The language of social and economic networks affords a first principle integration by simplifying the non-trivial concept of social and economic structure~\cite{martin_lee} to only dyadic relations.

As mentioned earlier, macroeconomic models are ill-suited to the study of collective phenomena at an intermediate level. Hence, alternative explanations of aggregate phenomena that match the expressiveness of economic models are required. Analytical sociology~\cite{ch1as_hdbk}, with its emphasis on explanation of collective emergent phenomena using mathematically formulated social mechanisms~\cite{ch2as_hdbk,ch11as_hdbk} anchored at the individual level, is an ideal candidate for this purpose. It is particularly powerful when combined with computational simulation, because when mathematically formulated models reach even a modest level of complexity, they often become analytically intractable. Simulation \textit{in silico} can yield approximate results for these more complex scenarios, which supplement the exact results reached by traditional methods.

Models constructed here, like those used in scientific inquiry in general, serve a specific purpose. In this work, we construct simple toy models to reproduce certain aspects of non-trivial wealth inequality distributions in the presence and absence of primitive forms of cooperation, clearly delineating the role of network structure in generating wealth inequality. We make no suggestions that these models \textit{explain} the phenomena of interest; we are only interested in constructing the simplest possible models, with no detailed empirical grounding, but with the potential to generate realistic looking inequality distributions. In doing so, we aspire to shed light on the \textit{true} social and economic mechanisms underlying the genesis and persistence of economic inequality, and network structure alone can produce large differences in economic outcomes.

\label{giniKindaSucks}
% SD: In light of recent discussions, we may want to downplay Gini and up-play the robustness of agents to exogenous shocks.
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures that enable economic interactions and cooperation mechanisms. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian-noise-driven linear dynamical systems, here with constant growth rate; the structure is modeled by networks, here with random graphs; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Although both the dynamical system and the network model are quite well understood, the precise interplay of structure and dynamics produces interesting emergent wealth distributions and robustness to scarcity. To the best of our knowledge, this specific combination of network structure and social dynamics has not been discussed in the computational social science (CSS) or mathematical sociology literature, and we consider this the primary contribution of this work.

Apart from enabling interactions between social actors, social structures like institutions also shape the form of cooperation and coordination mechanisms. The institutions can take the form of economic institutions, like banks and cooperatives; or the form of norms, like resource-sharing practices in societies. The original SriMil model~\cite{srimil}consists of a simple resource pooling arrangement where aggregates of agents pool their excess wealth in a common institution called ``proto-institutions,'' agreeing to provide this saved resource to individual agents in times of need. The model of resource pooling used in this work is identical to the SriMil model.

The analysis to be presented in later sections focuses only on homogeneous agents with simple drift-diffusion dynamics on Erd\H{o}s-R\'{e}nyi network models (ER); space constraints unfortunately prevent us from repeating the analysis on other standard \textit{textbook} networks like scale-free and small-world networks. We discuss empirical evidence for the role of social network structures in the concluding section of this paper, motivating the need for more expressive social network models.

In the next section, we discuss the mathematical formulation of the model. In section 3, we present the analysis of our simulation experiments, summarize key findings, and discuss why we chose the \texttt{julia} programming language~\cite{Julia-2017} for our implementation. In section 4, we discuss the limitations of our simple models, extensions to dynamics and networks more expressive than the ones presented here and planned future work.

Before discussing our model in greater mathematical detail, we discuss the model qualitatively, contrasting it with more familiar modeling approaches. The model presented here has much in common with models used in social-reality-inspired models in statistical physics~\cite{redner2001guide}, dynamic process models in network science~\cite{newman2018networks}, and computational social science and agent-based models~\cite{abm_review}; however, our modeling philosophy is somewhere at the interface of agent-based computational economics (ACE)~\cite{tesfatsion} and analytical sociology (AS)~\cite{ch1as_hdbk,ch2as_hdbk}. In ACE, we acknowledge its aspiration to develop bottom-up models of economic systems at all scales but restrain from its enthusiastic use of complex but well-calibrated detailed models of markets and agents~\cite{tesfatsion2017}. In AS, we assent its focus on social mechanisms in explaining social phenomena, but instead rely on simplified mechanisms with few parameters~\cite{ch11as_hdbk} with the specific goal of extracting insights from \textit{stylized} models.

@@

While both the stochastic models and ER models are well understood, proto formation dynamics, as coagulation processes in continuous time, remain to be better understood, mathematically. Also, while the mathematical analysis of dynamics of discrete-time stochastic models on networks is well studied, diffusion processes on networks seem under explored in the literature.

>>>>

\end{abstract}
\section{Introduction}
In recent years, the concerns and debates regarding wealth inequality and socioeconomic mobility have been one of the few unifying issues dominating the extremely polarized public spheres of the Global North. While economic and political inequality used to be discussed in heterodox economics circles, the contentious discussions on the topic within mainstream economics since the publication of Piketty's book~\cite{piketty2017capital} suggest a lack of consensus about basic foundational questions like the origins and persistence of economic inequality. Not surprisingly, traditional theories and tools of macro and micro economics are now being diagnosed for their limitations. Simultaneously, insights from related disciplines, along with novel models for scientific inquiry not usually associated with traditional econometrics, are being taken more seriously. The work presented here shares this spirit by integrating substantive ideas from anthropology, economic sociology~\cite{granovetter2017society} and urban sociology~\cite{sampson2012great} and using modeling approaches from computational social science~\cite{hedstrom2018}, computational economics~\cite{tesfatsion} and analytical sociology~\cite{hedstrom2011oxford} to explore the origins of inequality in a simple model of wealth dynamics in the presence of social structures.    
% SD: citation for "anthropology," the first element of the last sentence's list? Grief and/or White?

@@

The model presented here was inspired by our search for analogs of the grid world agent-based model (ABM) of Friesen and Mudigonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. This model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

The Friesen and Mudigonda model drew its inspiration from historical sociology, in Katz's influential study of middle of 19th century Hamilton, Canada~\cite{katz2013people}. Retaining this original motivation, we draw additional inspiration from economic sociology in the work of Granovetter~\cite{granovetter2005}; urban sociology, in the work of Sampson~\cite{sampson2002} and others; and in anthropology, in the work on cooperation in small to medium scale societies~\cite{avner1994,white2011kinship}, and others. The geographic and economic scale of the systems and the nature of social actors and time scale of interest in these different disciplineary approaches are all very different from the ones used to develop models of representative agents in macroeconomics~\cite{benhabib2018}, making the similarity between macroeconomic wealth dynamics models and our models not comparable without further justification. Elaborating on the interplay of conceptual and methodological ideas among these disciplines is beyond the scope of this article. Instead, we anchor this work in economic sociology and revisit the insights from the above disciplines in the article's concluding section.

The important role played by social structures in determining economic outcomes of individuals in a society is not in doubt~\cite{granovetter2005, jackson_rev2017}. Still, in the absence of a unifying foundation for sociology and economics, the full impact of this two-way interpenetration of economic and social structures is demonstrated only on a case-by-case basis. The language of social and economic networks affords a first principle integration by simplifying the non-trivial concept of social and economic structure~\cite{martin_lee} to only dyadic (binary) relations among actors\footnote{The concept social structures~\cite{martin_lee} is more general than the widely held belief in computational social science that social structure \textit{is} just a social network.}.

Models of wealth accumulation dynamics in both macro and micro economics typically assume the presence of markets and equilibria. Hence, such models are ill-suited to the study of collective phenomena at an intermediate level. Hence, alternative explanations of aggregate phenomena that match the expressiveness of economic models are required. Analytical sociology~\cite{ch1as_hdbk}, with its emphasis on explanation of collective emergent phenomena using mathematically formulated social mechanisms~\cite{ch2as_hdbk,ch11as_hdbk} anchored at the individual level, is an ideal candidate for this purpose. It is particularly powerful when combined with computational simulation, because when mathematically formulated models reach even a modest level of complexity, they often become analytically intractable. Simulation \textit{in silico} can yield approximate results for these more complex scenarios, which supplement the exact results reached by analytically tractable models.

Models constructed here, like those used in scientific inquiry in general, serve a specific purpose. In this work, we construct simple toy models to reproduce certain aspects of non-trivial wealth inequality distributions in the presence and absence of primitive forms of cooperation, clearly delineating the role of network structure in generating wealth inequality. We make no suggestions that these models \textit{explain} the phenomena of interest; we are only interested in constructing the simplest possible models, with no detailed empirical grounding, but with the potential to generate heterogeneous wealth outcomes reminiscent of ones obtained using more sophisticated models. In doing so, we aspire to shed light on the true social and economic mechanisms underlying the genesis and persistence of economic inequality, and network structure alone can produce large differences in economic outcomes.

\label{giniKindaSucks}
% SD: In light of recent discussions, we may want to downplay Gini and up-play the robustness of agents to exogenous shocks.
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures that enable economic interactions and cooperation mechanisms. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian-noise-driven linear dynamical systems, here with constant growth rate; the structure is modeled by networks, here with random graphs; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Just as with macroeconomic models, measures of wealth inequality like Gini coefficients are not so well suited for the non-market based wealth dynamics. We discuss this more in section 4.

Although both the dynamical system and the network model are quite well understood, the precise interplay of structure and dynamics produces interesting emergent wealth distributions and robustness to scarcity. To the best of our knowledge, this specific combination of network structure and social dynamics has not been discussed in the computational social science (CSS) or mathematical sociology literature, and we consider this the primary contribution of this work.

Apart from enabling interactions between social actors, social structures like institutions also shape the form of cooperation and coordination mechanisms. The institutions can take the form of economic institutions, like banks and cooperatives; or the form of norms, like resource-sharing practices in societies. The Friesen and Mudigonda model~\cite{srimil}consists of a simple resource pooling arrangement where aggregates of agents pool their excess wealth in a common institution called ``proto-institutions,'' agreeing to provide this saved resource to individual agents in times of need. The model of resource pooling used in this work is identical to this model. 
%% Add this in the acknowledgement section
The analysis to be presented in later sections focuses only on homogeneous agents with simple drift-diffusion dynamics on Erd\H{o}s-R\'{e}nyi network models (ER); space constraints unfortunately prevent us from repeating the analysis on other standard \textit{textbook} networks like scale-free and small-world networks. We discuss empirical evidence for the role of social network structures in the concluding section of this paper, motivating the need for more expressive social network models.

In the next section, we discuss the mathematical formulation of the model. In section 3, we present the analysis of our simulation experiments, summarize key findings, and discuss why we chose the \texttt{julia} programming language~\cite{Julia-2017} for our implementation. In section 5, we discuss the limitations of our simple models, extensions to dynamics and networks more expressive than the ones presented here and planned future work.

Before discussing our model in greater mathematical detail, we discuss the model qualitatively, contrasting it with more familiar modeling approaches. The model presented here has much in common with models used in social-reality-inspired models in statistical physics~\cite{redner2001guide}, dynamic process models in network science~\cite{newman2018networks}, computational social science, and agent-based models~\cite{abm_review}; however, our modeling philosophy is somewhere at the interface of agent-based computational economics (ACE)~\cite{tesfatsion} and analytical sociology (AS)~\cite{ch1as_hdbk,ch2as_hdbk}. In ACE, we acknowledge its aspiration to develop bottom-up models of economic systems at all scales but restrain from its enthusiastic use of complex but well-calibrated detailed models of markets and agents~\cite{tesfatsion2017}. In AS, we assent its focus on social mechanisms in explaining social phenomena, but instead rely on simplified mechanisms with few parameters~\cite{ch11as_hdbk} with the specific goal of extracting insights from \textit{stylized} models.

@@

While both the stochastic models and ER models are well understood, proto formation dynamics, as coagulation processes in continuous time, remain to be better understood, mathematically. Also, while the mathematical analysis of dynamics of discrete-time stochastic models on networks is well studied, diffusion processes on networks seem underexplored in the literature.
